---
title: "Machine Learning Assignment 1"
author: "Pamela Afful"
format: pdf
editor: visual
---

### Q1

```{r}

lq=function(x,q){
  coeff=2**q
  summands_store=c()
  for(k in 0:q) { # These are the ks we are summing over
  
   summands_k=(x**q)*(choose(q,k)*choose((q+k-1),q)  )
   summands_store=c(summands_store,summands_k)
  }
  
  f_x=coeff*sum(summands_store)
  
  return(f_x)
  
}



```

a\)

```{r}

# plot the above
# USE sapply to vectorize the calculation across a supplied vector
x_list=list(x=seq(-1,1,0.1))

x=seq(-1,1,0.1)
Qs=seq(1,5,1)

plot(NULL,xlim=c(-1,1),ylim=c(min(sapply(x,lq,q=5)),max(sapply(x,lq,q=5))),xlab='x',ylab='Lq(x)',main='Lq(x) Polynomials')
for(q_i in Qs){
  lines(x,sapply(x,lq,q=q_i),col='blue',lwd=1.2)

}


```

They are nonlinear curves at varying degrees

b\)

```{r}
set.seed(2)

alpha_qs=runif(3,-1,1)
beta_qs=runif(3,-1,1)

target_alpha=function(x,alpha_q,Qf){
  store_q=c()
  for(q_i in 0:Qf){
    
    store_q=c(store_q,alpha_q*(x**q_i))
  }
  
  ret=sum(store_q)
  return(ret)
}


target_beta=function(x,beta_q,Qf){
  store_q=c()
  for(q_i in 0:Qf){
    Lq=lq(x,q_i)
    store_q=c(store_q,beta_q*Lq)
  }
  
  ret=sum(store_q)
  return(ret)
}


# Plot the above target_alpha

plot(NULL,xlim=c(-1,1),ylim=c(-10,10),xlab='x',ylab='f(x)',main='Eqn(2) Random Target Functions')

for(alpha in alpha_qs){
  lines(x,sapply(x,target_alpha,alpha_q=alpha,Qf=5),col='blue',lwd=1.2)
}

# Plot the above target_beta


plot(NULL,xlim=c(-1,1),ylim=c(-100,100),xlab='x',ylab='f(x)',main='Eqn(2) Random Target Functions')

for(beta in beta_qs){
  lines(x,sapply(x,target_beta,beta_q=beta,Qf=5),col='red',lwd=1.2)
}


```

### Q2)

```{r}

q=10
Ns=seq(20,110,10)


sigmas=seq(0.2,1.1,0.1)



lq=function(x,q=10){ # q is the order
  coeff=2**q
  summands_store=c()
  for(k in 0:q) { # These are the ks we are summing over
  
   summands_k=(x**q)*(choose(q,k)*choose((q+k-1),q)  )
   summands_store=c(summands_store,summands_k)
  }
  
  f_x=coeff*sum(summands_store)
  
  return(f_x)
  
}


target_beta=function(x,beta_q,Qf,sig){
  epsi=rnorm(1,0,sig)
  store_q=c()
  for(q_i in 0:Qf){
    Lq=lq(x,q_i)
    store_q=c(store_q,beta_q*Lq)
  }
  
  ret=sum(store_q)+epsi
  return(ret)
}

set.seed(1)




#Generate a list of N datasets




n_sim=10

all_matrix=1:length(Ns)
for(sim in 1:n_sim){
  error_diffs=matrix(NA,nrow=length(sigmas),ncol=length(Ns),dimnames = list(sigmas,Ns))
for(sig_i in sigmas){
  for(n in Ns) {
  set.seed(1)
  beta_q=runif(1,-1,1)
  Qf=10
  x_sim=runif(n,-1,1)
  
  
  # apply target
  yi_sig_n=sapply(x_sim,target_beta,beta_q=beta_q,Qf=10,sig=sig_i)
  
  # fit h2 and h10
  
  h2=lm(yi_sig_n~poly(x_sim,2,raw=T))
  h10=lm(yi_sig_n~poly(x_sim,10,raw=T))
  
  
  # get oos predictions from above
  x_sim_oos= runif(n,-1,1)
  yi_oos=sapply(x_sim_oos,target_beta,beta_q=beta_q,Qf=10,sig=sig_i)
  h2_pred_n_sig=unname(predict(h2,data.frame(x_sim_oos)))
  h10_pred_n_sig=unname(predict(h10,data.frame(x_sim_oos)))
  
  # Get the errors
  # Using MSE
  h2_oos_err=mean((yi_oos-h2_pred_n_sig)**2)
  h10_oos_err=mean((yi_oos-h10_pred_n_sig)**2)
  
  # get diff
  i=which(Ns==n)
  j=which(sigmas==sig_i)
  error_diffs[i,j]=(h2_oos_err-h10_oos_err)
  

  
    
  }

  
}
  #concat all the matrices from n simulations of each N,sig combination
  all_matrix=rbind(error_diffs)
}

all_matrix[1,]

 # matrix of error differences
```

```{r}
c1=c(1,2,3)

for(i in 1:5){
  
   c1=rbind(c1,c(3,2,1))
}

c1
```

### Q3

```{r}
set.seed(2023)
f_x=function(x,sig=0){
  epsi=rnorm(1,0,sig)
  
  if(x<0){
    target=abs(x+1)-0.5
    return (target+epsi)
    
  }
  if(x>=0){
    target=abs(x-1)-0.5
    
    return(target+sig)
    
  }
}

# Simulte 5 time
n_sim=seq(1,5,1)
N=5

plot(NULL, axes = FALSE, xlim=c(-2,2),ylim=c(-0.5,0.5),xlab='xi',ylab='f(xi)',main='Target Function Simulation')
axis(1, pos = 0)
axis(2, pos = 0)

colours=c('green','blue','lightblue','orange','red')
for(i in seq_along(n_sim)){
  xi=sort(runif(N,-2,2))
  y=sapply(xi,f_x)
  #lines( y~xi,lwd=2,col=colours[1])
  points(y~xi,lwd=2,col=colours[1],pch=1)
  
  # Plot Hypotheses
  h1=lm(y~xi)
  h1_predict=unname(predict(h1,data.frame(xi)))
  #lines(h1_predict~xi,lwd=2,col=colours[3])
  points(h1_predict~xi,lwd=2,col=colours[3],pch=1)
  # 
  h2=lm(y~sin(pi*xi)+cos(pi*xi))
  h2_predict=unname(predict(h2,data.frame(xi)))
  #lines(h2_predict~xi,lwd=1,col=colours[4])
  points(h2_predict~xi,lwd=1,col=colours[4],pch=1)
  # 
  
}

legend("bottomright", 
       legend = c("f(x)", "H1", "H2"), # Text labels
       col = c(colours[1],colours[3],colours[4]),                           
       pch = 1, 
       cex=0.85,
       title = "Sample Plots",
       bty='n')  







```

```{r}
 h1
```

b\) Given that the underlying data generating process is non-linear, the non-linear hypothesis would be more suited to fit the data

c\)

```{r}
library(dplyr)
library(tidyverse)

# Funtion Arg
# f_tru: True DGP
# hyp: Hypothesis function
# sig: nr of std deviation from mean
# N: size of data
# nsim= nr of simulations/draws

oos_estimate=function(f_true,hyp,sig,N, nsim=1000){
  
  #RHS BV trade-off
  bias_store=c()
  prediction_sim_store=matrix(NA,nrow=N,ncol=nsim) # store predictions for each simulation in this matrix,col= prediction at simulation i
  pred_var_score=c()
  oos_err_store=c()
  for(i in 1:nsim){
      xi=sort(runif(N,-2,2)) # sample N samples
      y=sapply(xi,f_x) # apply the f_true
      
      h_predict=unname(predict(hyp,data.frame(xi))) # get prediction of the hypothesis on the sample
      
      # Calculate the bias at draw i
      bias_i= (y-h_predict)**2
      bias_store=c(bias_store,bias_i) # store the bias at i
      
      # Store predictions
      prediction_sim_store[,i]=h_predict # at each column, insert the prediction
      
      # get running mean of the prediction
      # use sapply for the first simulation draw
       mean_prediction=ifelse(i>1,apply(prediction_sim_store[,1:i],1,mean),sapply(prediction_sim_store[,1:i],mean)) 
       
      # how far each prediction is from the mean prediction
       pred_var_i=(h_predict-mean_prediction)**2
       pred_var_score=c(pred_var_score,pred_var_i)
       
       
    #LHS
       # create OOS datapoint
       xi_oos=sort(runif(N,-2,2)) # draw at random again
       y_oos=sapply(xi_oos,f_x)
       h_predict_oos=unname(predict(hyp,data.frame(xi_oos)))
       
       # calculate the error SE^2
       oos_err_i=(h_predict_oos-y_oos)**2
       oos_err_store=c(oos_err_store,oos_err_i)
       
       
      
   }
  
  #  get the averages
  model_bias2=round(mean(bias_store),4)
  model_var=round(mean(pred_var_score),4)
  both=round(model_bias2+model_var,4)
  oos_err_est=round(mean(oos_err_store),4)
  estimate_diff=round(abs(both-oos_err_est),4)
  
  
  return(list(model_bias2=model_bias2,model_var=model_var,both=both,oos_err_est=oos_err_est,estimate_diff=estimate_diff))
}


# Test the above
f_true=f_x
sig=0
N=5



h1_out=oos_estimate(f_true,h1,sig,N, nsim=1000);h1_out

h2_out=oos_estimate(f_true,h2,sig,N, nsim=1000);h2_out


rbind(data.frame(h1_out),data.frame(h2_out)) %>% 
  mutate(model=c('h1(x)','h2(x)')) %>% 
  column_to_rownames(var = "model")

```

```{r}

sigmas=seq(0,0.5,0.1)


f_true=f_x
N=5

# apply function for all sigmas
h1_sigma_function_values=sapply(sigmas,oos_estimate,f_true=f_x,hyp=h1,N=5,nsim=1000)
h2_sigma_function_values=sapply(sigmas,oos_estimate,f_true=f_x,hyp=h2,N=5,nsim=1000)

# create bias plot
plot(NULL,xlim=c(0,0.5),ylim=c(0,0.10),xlab='sigma',ylab='bias2',main=' Model Bias vs Noise Parameter (sigma)',)
lines(sigmas,unlist(h1_sigma_function_values[1,]),col='blue',lwd=1.5)
points(sigmas,unlist(h1_sigma_function_values[1,]),col='blue',pch = 1)
lines(sigmas,unlist(h2_sigma_function_values[1,]),col='red',lwd=1.5)
points(sigmas,unlist(h2_sigma_function_values[1,]),col='red',pch = 1)

legend("bottomright", 
       legend = c("h1(x)", "h2(x)"), # Text labels
       col = c('blue','red'),                           
       pch = 1, 
       cex=0.85,
       title = "Hypothesis Function",
       bty='n')  


plot(NULL,xlim=c(0,0.5),ylim=c(0,0.10),xlab='sigma',ylab='Variance',main=' Model Variance vs Noise Parameter (sigma)',)
lines(sigmas,unlist(h1_sigma_function_values[2,]),col='blue',lwd=1.5)
points(sigmas,unlist(h1_sigma_function_values[2,]),col='blue',pch = 1)
lines(sigmas,unlist(h2_sigma_function_values[2,]),col='red',lwd=1.5)
points(sigmas,unlist(h2_sigma_function_values[2,]),col='red',pch = 1)

legend("bottomright", 
       legend = c("h1(x)", "h2(x)"), # Text labels
       col = c('blue','red'),                           
       pch = 1, 
       cex=0.85,
       title = "Hypothesis Function",
       bty='n')  


plot(NULL,xlim=c(0,0.5),ylim=c(0,0.14),xlab='sigma',ylab='E[E_OOS_Err]',main=' Model Expected OOS Error vs Noise Parameter (sigma)',)
lines(sigmas,unlist(h1_sigma_function_values[4,]),col='blue',lwd=1.5)
points(sigmas,unlist(h1_sigma_function_values[4,]),col='blue',pch = 1)
lines(sigmas,unlist(h2_sigma_function_values[4,]),col='red',lwd=1.5)
points(sigmas,unlist(h2_sigma_function_values[4,]),col='red',pch = 1)

legend("bottomright", 
       legend = c("h1(x)", "h2(x)"), # Text labels
       col = c('blue','red'),                           
       pch = 1, 
       cex=0.85,
       title = "Hypothesis Function",
       bty='n')  



```

c\)

-   Linear Model has better OOS performance overall averaged across 1000 simulations and varying values of the noise parameter

-   Non linear model has lower bias but higer variance and the ooposite is observed for variance of the model

-   Given the estimated OOS Error, go with the linear model as it is more parsimonious

Q4

a\)

```{r}
dat = read.table('Collider_Data_2022.txt', h = TRUE,stringsAsFactors =T) 
dim(dat)

head(dat,10)
```

```{r}

plot(NULL, xlim=c(min(dat['X1']),max(dat['X1'])),ylim=c(min(dat['X2']),max(dat['X2'])), xlab='X1',ylab='X2',main='Classification Scatter Plot')

points(dat[dat$Yi1==1,]$X1,dat[dat$Yi1==1,]$X2,col='blue',pch=10)

points(dat[dat$Yi2==1,]$X1,dat[dat$Yi2==1,]$X2,col='red',pch=10)

points(dat[dat$Yi3==1,]$X1,dat[dat$Yi3==1,]$X2,col='purple',pch=10)

legend(x='topleft',
       legend = c("Yi1", 'Yi2','Yi3'),
       col=c('blue','red','purple'),
       pch=c(10,10,10),
       title = "Groups",
       cex=0.85,
       bty = "o")

```

-   Insert comment here

b\)

```{r}

soft_max=function(z){ # assume z is a matrix
  #z_ret=matrix(NA,nrow=dim(z)[1],ncol=dim(z)[2])
  num=exp(z)
  
  sums=colSums(num)
  result=t(as.matrix(t(t(num) / sums)))

  return(result)
  
}




```

```{r}

# g = matrix(c(
#   1.25,  0.83, -0.42,
#   2.18,  1.47,  0.95,
#   0.64,  2.31,  1.72,
#   1.53, -0.26,  0.38
# ), nrow = 4, ncol = 3)
# 
# g
# soft_max(g)
```

c\)

```{r}

#rm(list = ls())

X_feat=cbind(dat$X1,dat$X2)
Y=cbind(dat$Yi1,dat$Yi2,dat$Yi3)

choose_label=function(vector_prob){
  
  labels_vec=ifelse(vector_prob==max(vector_prob),1,0)
  return(labels_vec)
  
}



feedforward_nn=function(X_input,y_input,theta,m,nu,activation_fun_2=soft_max,activation_fun_1=tanh){
  
  # theta is the vector of initial weights
  #it is of length p*m+m*q+m+q (W1+W2+b1+b2)
  # in this code q=d2

  y=as.matrix(y_input)
  
  # Initialize relevant layer matrices
  
  
  N=dim(X_input)[1] # nr obs
  p=dim(X_input)[2] # nr of cols=nr of nodes in input layer for obs i
  #1. nr of nodes in input layer=2 features across N observations

  # Going into layer 1
  d0=p # nodes in input
  d1=m # nodes in hidden
  d2=dim(y)[2]

  X=t(as.matrix(X_input)) # obs as cols
  index_chunk1=p*m #W1
  index_chunk2=index_chunk1+1 #W2
  index_chunk3=index_chunk2+1 #B1
  index_chunk4=index_chunk3+1 #B2
  
  
  W1=matrix(theta[1:index_chunk1],nrow=d0,ncol=d1) # weight matrix from input layer to hidden layer
  #A1_mat=matrix(0,nrow=m,ncol=p)
  b1=matrix(theta[index_chunk3:((index_chunk3+m)-1)],nrow=m,ncol=1) # bias matrix going into hidden layer
  b1_ones_matrix=matrix(1,nrow=1,ncol=N)
  b1_mat=b1%*%b1_ones_matrix
# Going  into output/layer 2
  
  W2=matrix(theta[index_chunk2:((index_chunk2+(m*d2))-1)],nrow=d1,ncol=d2) # weight matrix from input layer to hidden layer
  
  b2=matrix(theta[index_chunk4:((index_chunk4+d2)-1)],nrow=d2,ncol=1)
  b2_ones_matrix=matrix(1,nrow=1,ncol=N) # Repeat this N times
  b2_mat=b2%*%b2_ones_matrix
  
 # Do one forward calc and get prediction
 
first_layer_comp=t(W1)%*%X+b1_mat

# apply activation
first_layer_comp=activation_fun_1(first_layer_comp)




#second_layer_cal

second_layer_comp=activation_fun_2(t(W2)%*%first_layer_comp+b2_mat)




#predictions across N observations 
# prediction = max probability
y_preds=second_layer_comp
y_pred_label=t(apply(y_preds,1,choose_label)) # one hot encoding



# control for very small values in the matrix
log_compute= function(mat, threshold = 1e-4) {
  ifelse(mat < threshold, log1p(mat), log(mat))
}



CE_mat=y*log_compute(y_preds) 
CE_error=-mean(apply(CE_mat,1,sum)) # normal non_penalized oobj

CE_error_L2=CE_error+(nu/N)*(sum(W1^2)+sum(W2^2)) # regularized obj

  
return(list(first_layer_comp = first_layer_comp,second_layer_comp = second_layer_comp, CE_error = CE_error, CE_error_L2 = CE_error_L2))
  
  
  
}

y_input=Y
X_input=X_feat

nu  = 0
m   = 10

p     = dim(X_input)[2] 
q     = dim(as.matrix(Y))[2]
npars = p*m+m*q+m+q
theta_rand = runif(npars,-1,1)


nu = 0
obj_pen = function(pars)
{
  res_model = feedforward_nn(X_input,y_input,pars,m,nu,activation_fun_2=soft_max,activation_fun_1=tanh)
  return(res_model$CE_error) 
}

obj_pen(theta_rand)

# Fit the neural network using a standard optimizer in R:
res_opt = nlm(obj_pen,theta_rand, iterlim = 1000)
res_opt
```

```{r}
set.seed(2)
N       = dim(X_feat)[1]
set     = sample(1:N,0.8*N,replace = FALSE)
X_train = X_feat[set,,drop = FALSE] #
Y_train = Y[set,,drop = FALSE]
X_val   = X_feat[-set,,drop = FALSE]
Y_val   = Y[-set,,drop = FALSE]

obj_pen = function(pars)
{
  res_model = feedforward_nn(X_input,y_input,pars,m,nu,activation_fun_2=soft_max,activation_fun_1=tanh)  
  return(res_model$CE_error_L2) # Reg Obj
}

n_nus = 15
nu_seq= exp(seq(-5,1,length = n_nus))
val_error = rep(0,n_nus)
for(i in 1:n_nus)
{
	nu = nu_seq[i] # Pick nu
	res_opt = nlm(obj_pen,theta_rand, iterlim = 1000) # Fit for that nu
	
	res_val      = feedforward_nn(X_input,y_input,theta_rand,m,res_opt$estimate,activation_fun_2=soft_max,activation_fun_1=tanh)
  val_error[i] = res_val$CE_error_L2
  
  print(paste0('Validation run:',i,' nu = ', nu))
  
  res_fitted = feedforward_nn(X_input,y_input,theta_rand,m,res_opt$estimate,activation_fun_2=soft_max,activation_fun_1=tanh)#neural_net(XX,YY,res_opt$estimate,m,nu)

}
```

```{r}

m=10
library(colorspace)
color.gradient = function(x, colors=c('magenta','white','lightblue'), colsteps=50)
{
  colpal = colorRampPalette(colors)
  return( colpal(colsteps)[ findInterval(x, seq(min(x),max(x), length=colsteps)) ] )
}


best_nu_index=which(val_error==min(val_error))

best_nu=n_nus[best_nu_index]
best_nu

# use this to fit model again 


obj_pen = function(pars)
{
  res_model =feedforward_nn(X_input,y_input,theta_rand,m,best_nu,activation_fun_2=soft_max,activation_fun_1=tanh)
  return(res_model$CE_error_L2) # Reg Obj
}

res_opt_best_nu = nlm(obj_pen,theta_rand, iterlim = 1000)


#Resp curve
cols=c('blue','red','purple')

M   = 5
x1  = seq(-4,+3,length = M)
x2  = seq(-4,+3,length = M)
xx1 = rep(x1,M)
xx2 = rep(x2,each = M)

plot(NULL, xlim=c(-4,3),ylim=c(-4,3),xlab='x1',ylab='x2')

abline(h= x2,v = x1, lwd = 0.3)

points(xx2~xx1,pch = 16, cex = 0.5)


XX = cbind(xx1,xx2)
YY = cbind(rep(1,M^2))
res_fitted=feedforward_nn(X_input,y_input,res_opt_best_nu$estimate,m,best_nu,activation_fun_2=soft_max,activation_fun_1=tanh) 

plot(XX[,2]~XX[,1],col = color.gradient(res_fitted$second_layer_comp ),pch = 16,cex = 0.5)
points(X_feat[,2]~X_feat[,1], col = cols[Y+1], pch = 16)


```
