---
title: "Machine Learning Assignment 1"
author: "Pamela Afful"
format: pdf
editor: visual
---

### Q1

```{r}

lq=function(x,q){
  coeff=2**q
  summands_store=c()
  for(k in 0:q) { # These are the ks we are summing over
  
   summands_k=(x**q)*(choose(q,k)*choose((q+k-1),q)  )
   summands_store=c(summands_store,summands_k)
  }
  
  f_x=coeff*sum(summands_store)
  
  return(f_x)
  
}



```

a\)

```{r}

# plot the above
# USE sapply to vectorize the calculation across a supplied vector
x_list=list(x=seq(-1,1,0.1))

x=seq(-1,1,0.1)
Qs=seq(1,5,1)

plot(NULL,xlim=c(-1,1),ylim=c(min(sapply(x,lq,q=5)),max(sapply(x,lq,q=5))),xlab='x',ylab='Lq(x)',main='Lq(x) Polynomials')
for(q_i in Qs){
  lines(x,sapply(x,lq,q=q_i),col='blue',lwd=1.2)

}


```

They are nonlinear curves at varying degrees

b\)

```{r}
set.seed(2)

alpha_qs=runif(3,-1,1)
beta_qs=runif(3,-1,1)

target_alpha=function(x,alpha_q,Qf){
  store_q=c()
  for(q_i in 0:Qf){
    
    store_q=c(store_q,alpha_q*(x**q_i))
  }
  
  ret=sum(store_q)
  return(ret)
}


target_beta=function(x,beta_q,Qf){
  store_q=c()
  for(q_i in 0:Qf){
    Lq=lq(x,q_i)
    store_q=c(store_q,beta_q*Lq)
  }
  
  ret=sum(store_q)
  return(ret)
}


# Plot the above target_alpha

plot(NULL,xlim=c(-1,1),ylim=c(-10,10),xlab='x',ylab='f(x)',main='Eqn(2) Random Target Functions')

for(alpha in alpha_qs){
  lines(x,sapply(x,target_alpha,alpha_q=alpha,Qf=5),col='blue',lwd=1.2)
}

# Plot the above target_beta


plot(NULL,xlim=c(-1,1),ylim=c(-100,100),xlab='x',ylab='f(x)',main='Eqn(2) Random Target Functions')

for(beta in beta_qs){
  lines(x,sapply(x,target_beta,beta_q=beta,Qf=5),col='red',lwd=1.2)
}


```

### Q2)

```{r}

q=10
Ns=seq(20,110,10)
sigmas=seq(0.2,1.1,0.1)
datasets=list()


lq=function(x,q=10){ # q is the order
  coeff=2**q
  summands_store=c()
  for(k in 0:q) { # These are the ks we are summing over
  
   summands_k=(x**q)*(choose(q,k)*choose((q+k-1),q)  )
   summands_store=c(summands_store,summands_k)
  }
  
  f_x=coeff*sum(summands_store)
  
  return(f_x)
  
}


target_beta=function(x,beta_q,Qf,sig){
  epsi=rnorm(1,0,sig)
  store_q=c()
  for(q_i in 0:Qf){
    Lq=lq(x,q_i)
    store_q=c(store_q,beta_q*Lq)
  }
  
  ret=sum(store_q)+epsi
  return(ret)
}

set.seed(1)



#Generate a list of N datasets

for(n in Ns){
  set.seed(1)
  beta_q=runif(1,-1,1)
  Qf=10
  x_sim=seq(-1,1,length.out=n)
  i=which(Ns==n)
  datasets[[i]]=sapply(x_sim,target_beta,beta_q=beta_q,Qf=10)
  
}


datasets

```

### Q3

```{r}
set.seed(2023)
f_x=function(x,sig=0){
  epsi=rnorm(1,0,sig)
  
  if(x<0){
    target=abs(x+1)-0.5
    return (target+epsi)
    
  }
  if(x>=0){
    target=abs(x-1)-0.5
    
    return(target+sig)
    
  }
}

# Simulte 5 time
n_sim=seq(1,5,1)
N=5

plot(NULL, axes = FALSE, xlim=c(-2,2),ylim=c(-0.5,0.5),xlab='xi',ylab='f(xi)',main='Target Function Simulation')
axis(1, pos = 0)
axis(2, pos = 0)

colours=c('green','blue','lightblue','orange','red')
for(i in seq_along(n_sim)){
  xi=sort(runif(N,-2,2))
  y=sapply(xi,f_x)
  #lines( y~xi,lwd=2,col=colours[1])
  points(y~xi,lwd=2,col=colours[1],pch=1)
  
  # Plot Hypotheses
  h1=lm(y~xi)
  h1_predict=unname(predict(h1,data.frame(xi)))
  #lines(h1_predict~xi,lwd=2,col=colours[3])
  points(h1_predict~xi,lwd=2,col=colours[3],pch=1)
  # 
  h2=lm(y~sin(pi*xi)+cos(pi*xi))
  h2_predict=unname(predict(h2,data.frame(xi)))
  #lines(h2_predict~xi,lwd=1,col=colours[4])
  points(h2_predict~xi,lwd=1,col=colours[4],pch=1)
  # 
  
}

legend("bottomright", 
       legend = c("f(x)", "H1", "H2"), # Text labels
       col = c(colours[1],colours[3],colours[4]),                           
       pch = 1, 
       cex=0.85,
       title = "Sample Plots",
       bty='n')  







```

```{r}
 h1
```

b\) Given that the underlying data generating process is non-linear, the non-linear hypothesis would be more suited to fit the data

c\)

```{r}
library(dplyr)
library(tidyverse)

# Funtion Arg
# f_tru: True DGP
# hyp: Hypothesis function
# sig: nr of std deviation from mean
# N: size of data
# nsim= nr of simulations/draws

oos_estimate=function(f_true,hyp,sig,N, nsim=1000){
  
  #RHS BV trade-off
  bias_store=c()
  prediction_sim_store=matrix(NA,nrow=N,ncol=nsim) # store predictions for each simulation in this matrix,col= prediction at simulation i
  pred_var_score=c()
  oos_err_store=c()
  for(i in 1:nsim){
      xi=sort(runif(N,-2,2)) # sample N samples
      y=sapply(xi,f_x) # apply the f_true
      
      h_predict=unname(predict(hyp,data.frame(xi))) # get prediction of the hypothesis on the sample
      
      # Calculate the bias at draw i
      bias_i= (y-h_predict)**2
      bias_store=c(bias_store,bias_i) # store the bias at i
      
      # Store predictions
      prediction_sim_store[,i]=h_predict # at each column, insert the prediction
      
      # get running mean of the prediction
      # use sapply for the first simulation draw
       mean_prediction=ifelse(i>1,apply(prediction_sim_store[,1:i],1,mean),sapply(prediction_sim_store[,1:i],mean)) 
       
      # how far each prediction is from the mean prediction
       pred_var_i=(h_predict-mean_prediction)**2
       pred_var_score=c(pred_var_score,pred_var_i)
       
       
    #LHS
       # create OOS datapoint
       xi_oos=sort(runif(N,-2,2)) # draw at random again
       y_oos=sapply(xi_oos,f_x)
       h_predict_oos=unname(predict(hyp,data.frame(xi_oos)))
       
       # calculate the error SE^2
       oos_err_i=(h_predict_oos-y_oos)**2
       oos_err_store=c(oos_err_store,oos_err_i)
       
       
      
   }
  
  #  get the averages
  model_bias2=round(mean(bias_store),4)
  model_var=round(mean(pred_var_score),4)
  both=round(model_bias2+model_var,4)
  oos_err_est=round(mean(oos_err_store),4)
  estimate_diff=round(abs(both-oos_err_est),4)
  
  
  return(list(model_bias2=model_bias2,model_var=model_var,both=both,oos_err_est=oos_err_est,estimate_diff=estimate_diff))
}


# Test the above
f_true=f_x
sig=0
N=5



h1_out=oos_estimate(f_true,h1,sig,N, nsim=1000);h1_out

h2_out=oos_estimate(f_true,h2,sig,N, nsim=1000);h2_out


rbind(data.frame(h1_out),data.frame(h2_out)) %>% 
  mutate(model=c('h1(x)','h2(x)')) %>% 
  column_to_rownames(var = "model")

```

```{r}

sigmas=seq(0,0.5,0.1)


f_true=f_x
N=5

# apply function for all sigmas
h1_sigma_function_values=sapply(sigmas,oos_estimate,f_true=f_x,hyp=h1,N=5,nsim=1000)
h2_sigma_function_values=sapply(sigmas,oos_estimate,f_true=f_x,hyp=h2,N=5,nsim=1000)

# create bias plot
plot(NULL,xlim=c(0,0.5),ylim=c(0,0.10),xlab='sigma',ylab='bias2',main=' Model Bias vs Noise Parameter (sigma)',)
lines(sigmas,unlist(h1_sigma_function_values[1,]),col='blue',lwd=1.5)
points(sigmas,unlist(h1_sigma_function_values[1,]),col='blue',pch = 1)
lines(sigmas,unlist(h2_sigma_function_values[1,]),col='red',lwd=1.5)
points(sigmas,unlist(h2_sigma_function_values[1,]),col='red',pch = 1)

legend("bottomright", 
       legend = c("h1(x)", "h2(x)"), # Text labels
       col = c('blue','red'),                           
       pch = 1, 
       cex=0.85,
       title = "Hypothesis Function",
       bty='n')  


plot(NULL,xlim=c(0,0.5),ylim=c(0,0.10),xlab='sigma',ylab='Variance',main=' Model Variance vs Noise Parameter (sigma)',)
lines(sigmas,unlist(h1_sigma_function_values[2,]),col='blue',lwd=1.5)
points(sigmas,unlist(h1_sigma_function_values[2,]),col='blue',pch = 1)
lines(sigmas,unlist(h2_sigma_function_values[2,]),col='red',lwd=1.5)
points(sigmas,unlist(h2_sigma_function_values[2,]),col='red',pch = 1)

legend("bottomright", 
       legend = c("h1(x)", "h2(x)"), # Text labels
       col = c('blue','red'),                           
       pch = 1, 
       cex=0.85,
       title = "Hypothesis Function",
       bty='n')  


plot(NULL,xlim=c(0,0.5),ylim=c(0,0.14),xlab='sigma',ylab='E[E_OOS_Err]',main=' Model Expected OOS Error vs Noise Parameter (sigma)',)
lines(sigmas,unlist(h1_sigma_function_values[4,]),col='blue',lwd=1.5)
points(sigmas,unlist(h1_sigma_function_values[4,]),col='blue',pch = 1)
lines(sigmas,unlist(h2_sigma_function_values[4,]),col='red',lwd=1.5)
points(sigmas,unlist(h2_sigma_function_values[4,]),col='red',pch = 1)

legend("bottomright", 
       legend = c("h1(x)", "h2(x)"), # Text labels
       col = c('blue','red'),                           
       pch = 1, 
       cex=0.85,
       title = "Hypothesis Function",
       bty='n')  



```

c\)

-   Linear Model has better OOS performance overall averaged across 1000 simulations and varying values of the noise parameter

-   Non linear model has lower bias but higer variance and the ooposite is observed for variance of the model

-   Given the estimated OOS Error, go with the linear model as it is more parsimonious

Q4

a\)

```{r}
dat = read.table('Collider_Data_2022.txt', h = TRUE,stringsAsFactors =T) 
dim(dat)

head(dat,10)
```

```{r}

plot(NULL, xlim=c(min(dat['X1']),max(dat['X1'])),ylim=c(min(dat['X2']),max(dat['X2'])), xlab='X1',ylab='X2',main='Classification Scatter Plot')

points(dat[dat$Yi1==1,]$X1,dat[dat$Yi1==1,]$X2,col='blue',pch=10)

points(dat[dat$Yi2==1,]$X1,dat[dat$Yi2==1,]$X2,col='red',pch=10)

points(dat[dat$Yi3==1,]$X1,dat[dat$Yi3==1,]$X2,col='purple',pch=10)

legend(x='topleft',
       legend = c("Yi1", 'Yi2','Yi3'),
       col=c('blue','red','purple'),
       pch=c(10,10,10),
       title = "Groups",
       cex=0.85,
       bty = "o")

```

-   Insert comment here

b\)

```{r}

soft_max=function(z){ # assume z is a vector
  #z_ret=matrix(NA,nrow=dim(z)[1],ncol=dim(z)[2])
  num=exp(z)
  
  sums=apply(num,2,sum)
  result <- t(as.matrix(t(t(num) / sums)))

  
  
  
#   for(i in 1:dim(num)[2] ){ # along cols
#     for(j in 1:dim(num)[1] ){ # along rows
#       
#     z_ret[j,i]=num[j,i]/apply(num,2,sum)[i]
#     
#   }
# }
# 
#   
  return(result)
#   
}

# Test
N=10

x=matrix(1:30,ncol=N,nrow=3)

dim(soft_max(x))

z=Y_initial


dim(as.matrix(t(soft_max(Y_initial))))
```

c\)

```{r}

#rm(list = ls())

X=cbind(dat$X1,dat$X1)
Y=cbind(dat$Yi1,dat$Yi2,dat$Yi3)


FeedForward = function(X, Y,iterlim = 1000, plt = FALSE)
{
  # Dimensions and initialize
  N = dim(X)[1]
  d =  dim(X)[2] # nr of features
  K=dim(Y)[2] # nr of classes we are classifying
  X = cbind(1,X) # add bias weight/term to the X
  #Y = matrix(T,ncol=1) # ensure that it stays a column matric
  W = matrix(0,d+1,K) #weight matrix for K classes, with nrow=d+1,ncol=nr_classes
  
  # start with initial prediction
  Y_initial=as.matrix(X%*%W)
  
  # Apply the activation
  Y_initial_pred=soft_max(Y_initial);
  
  Y_initial
  
}


```
